Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-08 paper/paper.tex
Abstract re-write

\begin{abstract}
This study examines the extent to which the outcomes of clinical trials can be predicted based on longitudinal properties of drug targets and diseases alone. We find that this is possible by comparing the historical performance of model-based target-disease pair prioritization methods to common baselines. Our primary objective is to demonstrate that statistical learning can effectively optimize such methods with no loss of interpretability. For example, non-negative linear models can produce simple weighting schemes across various types of human, animal and cell model evidence (for targets, diseases and pairings of the two) to identify target-disease pairs that advance beyond phase 2 trials with an average relative risk that is 2x higher than Open Targets composite scores. Other key characteristics of this study include: 1) a comprehensive longitudinal treatment of evidence as well as how it relates to leakage and reverse causality in biomedical research, 2) trial and/or drug details are not used in order to enable ranking for undeveloped targets/diseases, 3) analysis of the space of currently undeveloped, tractable targets with the highest likelihood of clinical success, 3) no data is used outside of Open Targets to ease reproduction and/or deployment, and 4) our method requires no expert knowledge and can easily support the inclusion of more lines of evidence over time, making it easy to operationalize.
% Note: OT multiplier (2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10
\end{abstract}

>>>>

\begin{abstract}

Selecting optimal drug targets for diseases represents a critical juncture in the creation of new medicines. In this study, we present an automated framework for selecting such targets through a synthesis of human, animal, and cell model evidence that features simple, highly interpretable statistical models. These models are optimized based on the historical clinical success of target-disease pairings, use only 33 predictors derived directly from individual lines of evidence in Open Targets and can predict clinical advancement for those pairings with a success rate that is 1.5-2x higher than heuristic, composite scores derived from the same evidence as well as common measures of genetic support that have already been established -- and observed in this study -- to confer a 2x higher likelihood of success. We find that this cumulative 3-4x advantage is maximized when also utilizing prior human clinical validation specific to targets and diseases independent of pairings, or other static properties of targets like conservation, expression specificity and gene essentiality. These predictors are rarely absent, which contrasts highly with the evidence for target-disease pairs that rarely exists among those that have ever reached clinical trials. The extent and non-random nature with which this absence manifests pose key challenges in the selection of performance measures and predictor derivation. In particular, we address the possibility for reverse causality between clinical outcomes and evidence discovery through the use of longitudinal predictors. This novel aspect of our work also enables us to directly evaluate which types of evidence are most likely to leak information about clinical outcomes when not temporalized. By formulating target selection as a machine learning problem in this manner, and after accounting for these unique challenges, this framework can be easily deployed to prioritize disease targets yet to undergo clinical development. We explore some of these present-day targets, with a focus on those that are tractable, and offer all code and data used in this analysis as a proof-of-concept for our more comprehensive, proprietary ranking platform.
% Note: OT multiplier (2x) taken from Sensitivity / Relative risk ratio distribution across configurations / mean@10
\end{abstract}
